---
title: "What drives social networks?\\linebreak[4]A gentle introduction to exponential random graph models (with a focus on \\color{USCCardinal} small networks\\color{black})"
#\\footnote{Contact: \\url{vegayon@usc.edu}. We thank members of our MURI research team, USC's Center for Applied Network Analysis, Andrew Slaughter, and attendees of the NASN 2018 conference for their comments.}"
author:
  - "\\textbf{George G Vega Yon}"
date: "LAERUG\\linebreak[4]June 10, 2019"
institute:
  - "\\includegraphics[width=.15\\linewidth]{usc.pdf}\\linebreak[4]Department of Preventive Medicine"
output:
  beamer_presentation:
    keep_tex: true
    includes:
      in_header: ergmitos-header.tex
# classoption: handout
fontsize: 10pt
aspectratio: 169
nocite: |
  @Csardi2015, @knitr, @rmarkdown, @R, @Handcock2006, @Wasserman1996, |
  @hunter2008, @Handcock2018, @vegayon2018, @vegayon2019
bibliography: bibliography.bib
handout: false
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(smallsize = function(before, options, envir) {
    if (before) {
        "\\footnotesize\n\n"
    } else {
        "\n\\normalsize\n\n"
    }
})
knitr::opts_chunk$set(echo = TRUE, smallsize=TRUE, out.width = ".6\\linewidth",
                      fig.align = "center")
```

## Social networks

\begin{figure}
\centering
\includegraphics[height=.7\textheight]{ukfaculty-igraph.png}
\caption{Friendship network of a UK university faculty. Source: \textbf{igraphdata} R package (Csardi, 2015). Figure drawn using the R package \textbf{netplot} (yours truly, https://github.com/usccana/netplot)}
\end{figure}

<!-- ## What drives \sout{\color{USCCardinal}social} networks? -->

<!-- If \color{gray}\textit{[blank]}\color{black}{} asks you to predict a network\pause -->

<!-- \Huge What kind of model?\pause -->

<!-- \Huge What features would you include?\pause -->

<!-- \normalsize -->

## What drives \sout{\color{USCCardinal}social} networks?

Why are you and I are \color{gray}\textit{[blank]} \color{black}? (friends, collaborators, etc.)\pause

```{r ergm-terms, echo=FALSE}
knitr::include_graphics("fig/friendly-terms.pdf")
```

\pause Let's build a model for this!

## Exponential Family Random Graph Models (ERGMs) 

We need to build a probability function for \includegraphics[width=.05\linewidth]{fig/g1.pdf}...\pause

\begin{centering}

\def\tbw{.6\linewidth}

\note{First, we will focus on counts: "How many edges?", "How many homophilic ties?".
  We will call them "sufficient statistics"}
  \note{As we do in \sout{life} statistics, let's assume it is an additive model (we add stuff up), in a weighted fashion (i.e. we have model parameters!)}
\note{And since we like things to be positive... we just exponentiate it!}
\note{Finally, as probabilities should add up to 1, we will divide the thing by the sum of all possible cases (the "normalizing constant")}
  
  \begin{tcolorbox}[width=\tbw]
  $\# edges, \#homophilic\;ties, \dots$
  \end{tcolorbox}\pause
  \begin{tcolorbox}[width=\tbw]
  $\theta_{1} \times \#edges + \theta_{2} \times \#homophilic\;ties + \dots$
  \end{tcolorbox}\pause
  \begin{tcolorbox}[width=\tbw]
  $\exp{\theta_{1} \times \#edges + \theta_{2} \times \#homophilic\;ties + \dots}$
  \end{tcolorbox}\pause
  \begin{tcolorbox}[width=\tbw]
  $\frac{\exp{\theta_{1} \times \#edges + \theta_{2} \times \#homophilic\;ties + \dots}}{\sum \exp{\dots}}$
  \end{tcolorbox}
  
\end{centering}

You got yourself an ERGM!

## ERGMs... the \textit{lingua franca} of SNA

<!-- - Seeks to answer the question: \emph{What local social structures gave origin to a given observed graph?}\pause -->

<!-- - The model is centered around a vector of \textbf{sufficient statistics} $\sufstats{}$, and is operationalized as: -->

\begin{figure}
\centering
  \includegraphics[width = .8\linewidth]{fig/parts-of-ergm.pdf}
\end{figure}


------

\centering

There is one problem with this model ... \linebreak[4]

\includegraphics[width = .5\linewidth]{fig/parts-of-ergm.pdf}\pause \linebreak[4]

\large because of \color[HTML]{af0000}$\GRAPH$\color{black},
 the \color[HTML]{5726e7} \textbf{normalizing constant}\color{black}{} is \linebreak[4] a summation of $2^{n(n-1)}$ terms \includegraphics[width=.05\linewidth]{fig/scared.png}!\normalsize

-----

To solve this, instead of directly computing this function, estimation is done by approximating ratios of likelihood functions instead (TL;DR we use simulations).

\begin{figure}
\includegraphics[width=.6\linewidth]{fig/simply-not.jpg}
\end{figure}

  <!-- \begin{equation} -->
  <!-- 	\Prcond{\Graph = \graph}{\params, \Indepvar} = \frac{% -->
  <!-- 		\exp{\transpose{\params}\sufstats{\graph, \Indepvar}}%	 -->
  <!-- 	}{ -->
  <!-- 		\sum_{\graph'\in\GRAPH}\exp{\transpose{\params}\sufstats{\graph', \Indepvar}} -->
  <!-- 	},\quad\forall \graph\in\GRAPH\label{eq:ergm} -->
  <!-- \end{equation} -->
  
  <!-- Where $\kappa\left(\params, \Indepvar\right)$ is the normalizing constant and equals $\sum_{\graph'\in\GRAPH}\exp{\transpose{\params}\sufstats{\graph', \Indepvar}}$. \pause -->
  
<!-- - The set of sufficient statistics reflects social and psychological mechanisms that are hypothesized to drive the network structure. Figure \autoref{fig:ergm-structs} shows some examples of values in $\sufstats{}$.\pause -->

<!-- - In the case of directed networks, $\GRAPH$ has $2^{n(n-1)}$ terms.\pause -->

<!-- - See Wasserman, Pattison, Robins, Snijders, Handcock, Butts, and others. -->


## Let's get going

We will use the famous Monk data from @sampson1969novitiate \pause

```{r ergm-monks, message=FALSE}
library(ergm)
data(samplk, package="ergm")
```
\pause
This is an object of class network

```{r ergm-monk-print}
samplk1
```


---

```{r ermg-vis, warning=FALSE, message=FALSE, out.width=".5\\linewidth"}
library(sna) # Tools for SNA
set.seed(1)  # Graph layout is usually random-driven
gplot(samplk1)
```

\pause Let's add some color and other features

---

```{r ergm-vis-cont, out.width=".5\\linewidth"}
set.seed(1)
cols <- viridisLite::magma(4)[as.factor((samplk1 %v% "group"))]
gplot(samplk1, vertex.cex = degree(samplk1)/4, vertex.col = cols, edge.col = "gray")
```


## A simple ergm model

*   Suppose we want to test wether homophily on \textit{group} (individuals of the same group tend to connect with each other) and transitive triads (the friend of my friend) are driving the structure: \pause

    ```{r}
    summary(samplk1 ~ edges + nodematch("group") + ttriad)
    ```
    
\pause

*   To estimate this model we do:
    
    ```{r ergm-mcmc1, message=FALSE, warning=FALSE, cache=TRUE}
    ans <- ergm(
      samplk1 ~ edges + nodematch("group") + ttriad,
      control = control.ergm(seed = 112)
      )
    
    ```

----

```{r ergm1-summary, echo=TRUE, cache=TRUE}
summary(ans)
```

----

Now its time for \LARGE small \Huge networks!\normalsize 

---

\begin{figure}
\includegraphics[height = \textheight]{fig/american-chopper-argument-ergmitos.png}
\end{figure}

## ERGMs for Small Networks
			
-   In the case of small networks (e.g. at most 6 nodes), the calculation of \color[HTML]{5726e7} \textbf{normalizing constant}\color{black}{} becomes computationally feasible.\pause

-   This allows direct calculation of the likelihood, \textbf{avoiding the need for simulations} and allowing us to obtain Maximum Likelihood Estimates using \textit{standard} optimization techniques.\pause

-   In addition, most of the time samples of small networks include multiple of them, e.g.: Families, Small teams (like our data), Ego-nets, etc.\pause

-   This makes pooled ERGM estimates a natural way of modeling the data.\pause

-   This and more can be found in the [\textbf{ergmito}](https://github.com/muriteams/ergmito) R package \href{https://github.com/muriteams/ergmito}{(\includegraphics[width=.025\linewidth]{fig/github.png}/muriteams/ergmito)}

---

Sidetrack...

\begin{minipage}[c]{1\linewidth}
\large \textbf{ito, ita}: From the latin -\textit{\=ittus}. suffix in Spanish used to denote small or affection. e.g.:

\hspace{.5cm} \textit{¡Qué lindo ese perr\textcolor{USCCardinal}{\textbf{ito}}!} / \textit{What a beautiful little dog!}

\hspace{.5cm} \textit{¿Me darías una tac\textcolor{USCCardinal}{\textbf{ita}} de azúcar?} / \textit{Would you give me a small cup of sugar?}
\normalsize
\end{minipage}\pause

\alert{Special thanks to George Barnett who proposed the name during the 2018 NASN!}

## `ergmito` example

```{r loading-fivenets, cache=TRUE}
library(ergmito)
data(fivenets, package = "ergmito")
```

```{r plotfivenets, warning=FALSE, message=FALSE, echo=FALSE, fig.width=6, fig.height=3, out.width='.5\\linewidth', fig.align='center', cache=TRUE}
library(sna)
library(network)
op <- par(mfrow = c(2, 3), mai=rep(0, 4), oma = rep(0, 4))
USCCARDINAL <- rgb(153, 0, 0, maxColorValue = 255)
ans <- lapply(fivenets, function(f) {
  gplot(
    f,
    vertex.cex = 2,
    vertex.col = c("white", USCCARDINAL)[
      get.vertex.attribute(f, "female") + 1
    ]
    )
  })
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
legend("center", fill = c("white", USCCARDINAL), legend = c("Male", "Female"), cex=1, bty="n")
par(op)
```

----

```{r fivenets-1, cache=TRUE}
# Looking at one of the five networks
fivenets[[1]]
```

\pause How can we fit an ERGMito to this 5 networks?

## `ergmito` example (cont'd)

The same as you would do with the `ergm` package:\linebreak[4]

```{r fit-fivenets, cache=TRUE}
model1 <- ergmito(fivenets ~ edges + nodematch("female"))
summary(model1)
```

---

Some features of this (\includegraphics[width=.1\linewidth]{lifecycle-experimental-orange.pdf}) R package

\begincols

\begincol{.65\linewidth}
\begin{itemize}
  \setlength\itemsep{.5cm}
\item<2-> Built on top of \href{https://statnet.org}{\textbf{statnet}}'s \href{https://github.com/statnet/ergm}{\texttt{ergm}} R package.

\item<3-> Allows estimating ERGMs for small networks (less than 7 and perhaps 6) via MLE.

\item<4-> Implements pooled ERGM models.

\item<5-> Includes a simulation function for efficiently drawing samples of small networks, and by \textbf{efficiently} we mean \textbf{fast}.
\end{itemize}

\endcol

\begincol{.3\linewidth}

\includegraphics{fig/logo.png}

\endcol

\endcols

\uncover<6->{And much more!}

---

## Thanks!

\begin{centering}
\includegraphics[width = .4\linewidth]{fig/the-end.jpg}

\large \textbf{\textcolor{USCCardinal}{George G. Vega Yon}}


\href{mailto:vegayon@usc.edu}{vegayon@usc.edu} 

\href{https://ggvy.cl}{https://ggvy.cl} 

\includegraphics[width=.02\linewidth]{github.png}\href{https://github.com/gvegayon}{gvegayon}
\includegraphics[width=.02\linewidth]{twitter.png}\href{https://twitter.com/gvegayon}{gvegayon}

\end{centering}


---

\appendix

## Structures

\def\fig1width{.45\linewidth}
\begin{figure}
\centering
\begin{tabular}{m{.2\linewidth}<\centering m{.4\linewidth}<\raggedright}
\toprule Representation & Description  \\ \midrule
\includegraphics[width=\fig1width]{terms/mutual.pdf} & Mutual Ties (Reciprocity)\linebreak[4]$\sum_{i\neq j}y_{ij}y_{ji}$  \\
\includegraphics[width=\fig1width]{terms/ttriad.pdf} & Transitive Triad (Balance)\linebreak[4]$\sum_{i\neq j\neq k}y_{ij}y_{jk}y_{ik}$  \\
\includegraphics[width=\fig1width]{terms/homophily.pdf} & Homophily\linebreak[4]$\sum_{i\neq j}y_{ij}\mathbf{1}\left(x_i=x_j\right)$ \\
\includegraphics[width=\fig1width]{terms/nodeicov.pdf} & Covariate Effect for Incoming Ties\linebreak[4]$\sum_{i\neq j}y_{ij}x_j$ \\
\includegraphics[width=\fig1width]{terms/fourcycle.pdf} & Four Cycle\linebreak[4]$\sum_{i\neq j \neq k \neq l}y_{ij}y_{jk}y_{kl}y_{li}$  \\
\bottomrule
\end{tabular}
\caption{\label{fig:ergm-structs}Besides of the common edge count statistic (number of ties in a graph), ERGMs allow measuring other more complex structures that can be captured as sufficient statistics. }
\end{figure}

## References {.allowframebreaks}
