---
title: 'Intro to R (part 2):<br>R language fundamentals<br><img src="fig/trojan-rlogo.svg" style="width:250px;">'
author: George G. Vega Yon<br>University of Southern California<br>Department of Preventive Medicine
date: "<br>USC Integrative Methods of Analysis for Genomic Epidemiology (IMAGE)<br>Department of Preventive Medicine<br>July 7th, 2018"
bibliography: intro2r.bib
---

# Creating objects

*   In R you can create new objects by either using the assign operator (`<-`) or the equal sign `=`, for example, the following 2 are equivalent:
    
    ```{r assignment-1, collapse=TRUE, comment=""}
    a <- 1
    a =  1
    ```
    
    And it works for any type of data
    
    ```{r assignment-2, collapse=TRUE, comment=""}
    # Hey! here is a comment!.
    # Comments in R can be either at the begining of a line ...
    b <- "aloha" # ... or at the end of it!
    ```
    
    Historically the assign operator is the most common used.
    
----

# Data types 

*   The most basic data (object) type in R are vectors.
    
    ```{r}
    myvec1 <- vector("logical", 3)  # This is a logical vector of length 3
    myvec2 <- c(FALSE, TRUE, FALSE) # This is ALSO a logical vector of length 3
    myvec2
    length(myvec2)
    ```
      

*   Vectors can be of different modes: `logical`, `integer`, `numeric` (double), `complex`, `character`, and `raw`.
    (so one mode per vector!).
    
*   Vectors can also be lists, which is a VERY special type of object in R (most of its objects are lists!)
    ```{r}
    mylist1 <- vector("list", 2) # This is a list with 2 empty slots
    mylist2 <- list(NULL, NULL)  # This is ALSO a list with 2 empty slots
    mylist1
    ```

----

*   Moreover, lists elements can have names
    
    ```{r}
    mylist2 <- list(first = NULL, second = 1, third = c("a", "b", "c"))
    mylist2
    ```
    
    And we can access them as follows
    
    ```r
    # Getting the 3rd element of the list
    mylist2$third      # By name using the dollar sign
    mylist2[[3]]       # By position
    mylist2[["third"]] # By name using a string
    ```
    
    ```{r, echo=FALSE}
    mylist2[["third"]]
    ```


# Attributes and Structure

*   Objects in R have attributes (metadata)

    ```{r}
    # Here are the attributes of the iris dataset
    attributes(iris)    
    ```

    ```{r}
    head(iris)
    ```
    

----

*   You can take a general look at the structure of an R object with the `str()` function
    ```{r}
    # Let's take a look at the iris dataset, again
    str(iris)
    ```

# Missing values {.small}

R has different types of Missing values:

-   `NA`: no information, has length 1,
-   `NULL`: which has length 0,
-   `Inf`: Infinite, and
-   `NaN`: Not a Number
        
```{r missings, collapse=TRUE}
str(c(NA, 1L))        # Integers can have NAs
str(c(NaN, 1L, Inf))  # But not NaN or Inf (automatic coercion)
str(c(-Inf, 1, NULL, +Inf)) # And Nulls are... of length 0!
```

*   In the second line, we have an example of *automatic coercion*

*   These have companion functions `is.na()`, `is.null`, `is.infinite` (or `is.finite()`, which covers NA, Inf, and NaN), and `is.nan`.

# Questions

1.  What is the mode of the following vector `myvector <- c(NA, NaN, Inf)`? (try not to use the `mode()` function in R)

2.  The `c()` function can be used with other vectors, for example

    ```{r}
    my_integer_vector <- c(1L, 2L, 3L)    
    my_string_vector  <- c("hello", "world")
    ```
    
    What is the mode of the vector `c(my_integer_vector, my_string_vector)`?
    
3.  What do each one of the functions `is.na()`, `is.null`, `is.infinite`, `is.finite()`, and `is.nan` return on the vector `myvector`?

    ```{r testing-is, eval=FALSE, echo=FALSE}
    is.na(myvector)
    is.infinite(myvector)
    is.finite(myvector)
    is.nan(myvector)
    ```

4.    What are the attributes of the following object `mymat <- matrix(runif(12), ncol=4)`?

# Linear Algebra

*   **Matrix**: A special class of array, are vectors with a `dim` attribute of length 2 (number of rows and number of columns).
    
    ```{r}
    mymat <- matrix(1:9, ncol=3) # A matrix with numbers 1 through 9 with 3 columns
    mymat
    str(mymat)
    ```
    
    R stores elements in column major-order. We can access matrices elements in the following way.
    
    ```{r, collapse=TRUE}
    mymat[1, 2] # The element in the (1, 2) location of the matrix
    mymat[4]    # The fourth element in the vector... which is (1, 2)
    ```
    
    Matrices can have row and column names too.
  
----

*   Some matrix fundamental operators, for matrices `A`, and `B` (both square) with the same dimensions (just for the examples):

    *   Element wise product, addition, substraction, and quotient: `A * B`, `A + B`, `A - B`, `A/B`
    *   Transpose `t(A)`
    *   Inverse `solve(A)`
    *   Column and row bind, `cbind(A, B)`, `rbind(A, B)`
    
    See the `help("%*%")`, `help("cbind")`

# Questions

*   Given the following matrices `A`, `B`, and C
    
    ```{r, R.options=list(digits=2)}
    set.seed(122)
    A <- matrix(rnorm(12), ncol=3)
    B <- matrix(rnorm(12), nrow=3)
    C <- A %*% cbind(1:3) + rnorm(4) # Adding a vector of length 4!
    ```
    
    Compute
    
    1.  Matrix multiplication
    2.  Transpose of A
    3.  Element-wise product of A and B'
    4.  The inverse of the cross product of A
    5.  $(A^\mbox{t} A)^{-1} A^\mbox{t} C$
  

    ```{r, eval=FALSE, echo=FALSE}
    # Matrix multiplication:
    A %*% B

    # Transpose
    t(A)
    
    # Element-wise product
    A*t(B)

    # Inverse
    solve(t(A) %*% A)
    
    # OLS
    solve(t(A) %*% A) %*% t(A) %*% C
    ```

# Other fundamental types

Other fundamental type of objects in R are:

*   **Factors**: A set of *integer codes* with labels (levels).
    
    ```{r}
    myfactor <- factor(c("a", "a", "b", "xyz"))
    myfactor
    ```
    
    We treat factors as vectors.

----

*   **Data frames**: "A data frame is a list of variables of the same number of rows with unique row names, given class "data.frame"."
    
    ```{r}
    # Data frames can have multiple type objects, here, factor and numeric
    mydf <- data.frame(a_letter = c("a", "b", "c"), a_number = c(1, 2, 100))
    mydf
    str(mydf)
    ```
    
    We can access data frame objects as follow
    
    ```r
    mydf$a_letter      # Dollar sign
    mydf[[1]]          # Position of the column
    mydf[["a_letter"]] # Name of the column
    ```
    
    ```{r echo=FALSE}
    mydf[["a_letter"]]
    ```


# Statistical Functions

*   R has a very nice set of families of distributions. In general, distribution functions have the following name structure:
    
    a.  **Random Number Generation**: `r[name-of-the-distribution]`, e.g. `rnorm` for normal, `runif` for uniform.
    
    b.  **Density function**: `d[name-of-the-distribution]`, e.g. `dnorm` for normal, `dunif` for uniform.
    
    c.  **Cumulative Distribution Function (CDF)**: `p[name-of-the-distribution]`, e.g. `pnorm` for normal, `punif` for uniform.
    
    d.  **Inverse of the CDF (quantile)**: `q[name-of-the-distribution]`, e.g. `qnorm` for the normal, `qunif` for the uniform.
    
*   In the case of [**pseudo random numbers**](https://en.wikipedia.org/wiki/Pseudorandom_number_generator){target="_blank"} (the `r` prefix), it is important to **always** set the seed to ensure **reproducibility**

    ```{r setseed, collapse=TRUE}
    # First run
    set.seed(12)
    rnorm(4)
    
    # Second run
    set.seed(12)
    rnorm(4)
    ```
    
    More distributions available at `??Distributions`.

---

*   Take a look at the normal distribution

```{r}
set.seed(12)
op <- par(mfrow = c(2,2))
hist(rnorm(1e5))
curve(qnorm)
curve(pnorm, xlim=c(-3, 3))
curve(dnorm, xlim=c(-3, 3))
par(op)
```

---

*   Take a look at the exponential

```{r}
set.seed(12)
op <- par(mfrow = c(2,2))
hist(rexp(1e5))
curve(qexp)
curve(pexp, xlim=c(0, 6))
curve(dexp, xlim=c(0, 6))
par(op)
```

# Questions

1.  Draw 1e5 samples from a chi2 with 2 degrees of freedom (hint: check `?Distributions`).

```{r, echo=FALSE, eval=FALSE}
set.seed(778)
Z <- rchisq(1e5, 2)
hist(Z, breaks=100)
```
    

2.  Draw 1e5 samples from a chi2 with 2 degrees of freedom using `rnorm` (hint: Recall that if $X\sim N(0,1)$, then $X^2\sim\chi^2_1$, and if $X, Y\sim N(0,1)$, then $X^2 + Y^2\sim\chi^2_2$).

```{r, echo=FALSE, eval=FALSE}
set.seed(778)
Z <- rnorm(1e5)^2 + rnorm(1e5)^2
hist(Z, breaks=100)
```

#  Control-flow statements

*   For-loops

    ```{r control-flow}
    # The oldfashion forloop
    for (i in 1:10) {
      print(paste("I'm step", i, "/", 10))
    }
    ```
    
---

*   Ifelse

    ```{r control-flow2}
    # A nice ifelse
    for (i in 1:10) {
      
      if (i %% 2) # Modulus operand
        print(paste("I'm step", i, "/", 10, "(and I'm odd)"))
      else
        print(paste("I'm step", i, "/", 10, "(and I'm even)"))
      
    }
    ```
    
---

*   While

    ```{r control-flow3}
    
    # A while
    i <- 10
    while (i > 0) {
      print(paste("I'm step", i, "/", 10))
      i <- i - 1
    }
    ```

    
# Functions

*   Understanding functions, and functional programming can be very useful for your day-to-day work with R (and any programming language).

*   Functions can be created in a single line, or span multiple lines of code.

*   Usually, functions are structured as follow:
    
    ```r
    name_of_the_function <- function(
      argument1,    # First argument, required value
      argument2 = 1 # Second argument, has a default (so is optional)
    ) {
    
      [your neat R code here]
      
      return(answer)
    
    }
    
    ```

---

*   This is the very basic that we can write in R
        
    ```{r}
    # A function that does nothing
    my_lame_function <- function() {}
    my_lame_function()
    ```

---

*   Here is a more complicated function
    
    ```{r}
    calc_sum <- function(A, B) {
      A + B
    }
    
    calc_sum(1, 5)
    ```

---

*   We can use recursions too!

    ```{r}
    # An inefficient implementation of the Fibonacci number
    fib <- function(i) {
      
      if (i <= 2)
        return(1)
      
      fib(i - 2) + fib(i - 1)
      
    }
    
    c(fib(1), fib(2), fib(4), fib(5), fib(6), fib(7), fib(8))
    ```

